---
title: 'Supplement 3'
subtitle: "Fraction of oxygenated haemoglobin (FO2Hb)"
author: "Peter Kamerman and Etienne Basson"
date: "Last updated: `r format(Sys.Date(), '%d %B %Y')`"
mainfont: Helvetica
geometry: margin=15mm
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(magrittr)
library(skimr)
library(lmerTest)
library(merTools)
library(sjPlot)
library(lqmm)
library(MuMIn)
library(knitr)
library(patchwork)

# ggplot theme
theme_set(new = theme_bw(base_size = 18) +
              theme(panel.grid = element_blank(),
                    plot.title = element_text(size = 18),
                    plot.subtitle = element_text(size = 14)))

# Create figures directory
if(!dir.exists('figures')){
    dir.create('figures')
}

# skimr functions
my_skim <- skim_with(numeric = sfl(hist = NULL))

# Set chunk options
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      fig.path = 'figures/supplement-3-FO2Hb/',
                      fig.height = 5.8)
```

----

# Research question

1. Determine whether there is an association between blood colour components (CIE L\*a\*b\*) and fraction of oxygenated haemoglobin.

2. Generate a predictive model of the (CIE L\*a\*b\*) colour component from fraction of oxygenated haemoglobin.

----

# Load data 

```{r load_data}
data <- read_rds('data-cleaned/clean-data.rds')
```

----

# Quick look at the data

```{r quick_look}
dim(data)
names(data)
glimpse(data)
data[, !(names(data) %in% c('Trial'))] %>% 
    select_if(is.numeric) %>% 
    my_skim() 
```

----

# Prepare data

```{r prepare_data}
# Convert Animal ID to a factor
data %<>%
    mutate(Animal_ID = factor(Animal_ID,
                              levels = c('I1', 'I2', 'I3', 'I4', 'I5',
                                         'I6', 'I8', 'I9', 'I10', 'I11')))
```

----

# L*

## Exploratory plots

### Linear model

```{r exploratory_lstarA}
# Composite plot
ggplot(data = data) +
    aes(y = l_star,
        x = O2Hb,
        fill = Animal_ID) +
    geom_point(shape = 21,
               size = 3) +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm',
                fill = 'blue') +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs L*',
         subtitle = 'All animals together\nRegression line is simple linear model',
        x = 'FO2Hb (%)',
        y = 'L*') 
```

```{r exploratory_lstarB, fig.height = 8}
# Small multiples
ggplot(data = data) +
    aes(y = l_star,
        x = O2Hb) +
    geom_point(shape = 21,
               size = 3, 
               fill = '#FFFFFF') +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm') +
    labs(title = 'FO2Hb vs L*',
         subtitle = 'Facetted by animal\nRegression line: linear',
         y = 'L*',
         x = 'FO2Hb (%)') +
    facet_wrap(~ Animal_ID, 
               ncol = 3)

```

### Quadratic model

```{r exploratory_lstarC}
# Composite plot
ggplot(data = data) +
    aes(y = l_star,
        x = O2Hb,
        fill = Animal_ID) +
    geom_point(shape = 21,
               size = 3) +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm',
                fill = 'blue',
                formula = y ~ poly(x, 2)) +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs L*',
         subtitle = 'All animals together\nRegression line is a quadratic polynomial model',
        x = 'FO2Hb (%)',
        y = 'L*') 
```

```{r exploratory_lstarD, fig.height = 8}
# Small multiples
ggplot(data = data) +
    aes(y = l_star,
        x = O2Hb) +
    geom_point(shape = 21,
               size = 3, 
               fill = '#FFFFFF') +
    geom_smooth(se = FALSE,
                size = 0.8,
                colour = 'blue',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    labs(title = 'FO2Hb vs L*',
         subtitle = 'Facetted by animal\nRegression line: quadratic polynomial',
         y = 'L*',
         x = 'FO2Hb (%)') +
    facet_wrap(~ Animal_ID, 
               ncol = 3)
```

## Model selection

Model selection based on the outcome of likelihood ratio test of goodness of fit.

```{r model_lstar}
# Null model
lstar_null <- lmer(l_star ~ 1 + (1 | Animal_ID),
                   data = data,
                   REML = FALSE)

## First-order polynomial (linear) model
lstar_linear <- lmer(l_star ~ O2Hb + (1 | Animal_ID),
                     data = data,
                     REML = FALSE)

## Second-order polynomial (quadratic) model
lstar_poly <- lmer(l_star ~ poly(O2Hb, 2) + (1 | Animal_ID),
                   data = data,
                   REML = FALSE)

# Compare model fits
anova(lstar_null, lstar_linear, lstar_poly, test = 'LRT')
```

Best fit model is the second-order polynomial (quadratic) model.

## Explore the best fit model

```{r lstar_poly}
# Summary output
summary(lstar_poly)

# Check model assumptions
plot_model(model = lstar_poly,
           type = 'diag')[2:4]

# R-squared (based on likelihood ratio test)
r.squaredLR(lstar_poly, null = lstar_null)

# Root mean square error
RMSE.merMod(lstar_poly)
```

## Generate prediction intervals

```{r lstar_interval}
# Prediction interval
lstar_lqmm <- lqmm(fixed = l_star ~ poly(O2Hb, 2),
                   random = ~ 1,
                   group = ~ Animal_ID,
                   tau = c(0.025, 0.976),
                   data = data)

lstar_interval <- as.data.frame(predict(lstar_lqmm, level = 0))

names(lstar_interval) <- paste0('Q', c(2.5, 97.5))

lstar_interval <- cbind(data, lstar_interval)

# Numeric output
lstar_interval %>% 
    dplyr::select(O2Hb, l_star, Q2.5, Q97.5) %>% 
    arrange(O2Hb) %>% 
    unique(.) %>% 
    kable(caption = 'L*: 95% prediction interval',
          col.names = c('FO2Hb (%)', 'L*', 'Q2.5', 'Q97.5'))

# Plot output
ggplot(data = lstar_interval) + 
    aes(x = O2Hb) +
    geom_point(data = data,
               aes(y = l_star,
                   x = O2Hb,
                   fill = Animal_ID),
               shape = 21,
               size = 3) +
    geom_smooth(data = lstar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = lstar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs L*',
         subtitle = '95% prediction interval',
         y = 'L*',
         x = 'FO2Hb (%)')
```

## Generate predicted values 

```{r lstar_predicted}
# New data
lstar_grid <- expand.grid(O2Hb = c(15, 25, 35, 45, 55, 65, 75, 85, 95),
                          Animal_ID = 'I12')

# Generate predicted vs independent variable O2Hb
lstar_predicted <- predict(lstar_poly, 
                           lstar_grid, 
                           allow.new.levels = TRUE)

# Tabulate
data.frame(O2Hb = lstar_grid$O2Hb,
           lstar = lstar_predicted) %>% 
    kable(caption = 'L* predicted values',
          col.names = c('FO2Hb (%)', 'L*'))
```

----

# a*

## Exploratory plots

### Linear model

```{r exploratory_astarA}
# Composite plot
ggplot(data = data) +
    aes(y = a_star,
        x = O2Hb,
        fill = Animal_ID) +
    geom_point(shape = 21,
               size = 3) +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm',
                fill = 'blue') +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs a*',
         subtitle = 'All animals together\nRegression line is simple linear model',
        x = 'FO2Hb (%)',
        y = 'a*') 
```

```{r exploratory_astarB, fig.height = 8}
# Small multiples
ggplot(data = data) +
    aes(y = a_star,
        x = O2Hb) +
    geom_point(shape = 21,
               size = 3, 
               fill = '#FFFFFF') +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm') +
    labs(title = 'FO2Hb vs a*',
         subtitle = 'Facetted by animal\nRegression line: linear',
         y = 'a*',
         x = 'FO2Hb (%)') +
    facet_wrap(~ Animal_ID, 
               ncol = 3)

```

### Quadratic model

```{r exploratory_astarC}
# Composite plot
ggplot(data = data) +
    aes(y = a_star,
        x = O2Hb,
        fill = Animal_ID) +
    geom_point(shape = 21,
               size = 3) +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm',
                fill = 'blue',
                formula = y ~ poly(x, 2)) +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs a*',
         subtitle = 'All animals together\nRegression line is a quadratic polynomial model',
        x = 'FO2Hb (%)',
        y = 'a*') 
```

```{r exploratory_astarD, fig.height = 8}
# Small multiples
ggplot(data = data) +
    aes(y = a_star,
        x = O2Hb) +
    geom_point(shape = 21,
               size = 3, 
               fill = '#FFFFFF') +
    geom_smooth(se = FALSE,
                size = 0.8,
                colour = 'blue',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    labs(title = 'FO2Hb vs a*',
         subtitle = 'Facetted by animal\nRegression line: quadratic polynomial',
         y = 'a*',
         x = 'FO2Hb (%)') +
    facet_wrap(~ Animal_ID, 
               ncol = 3)
```

## Model selection

Model selection based on the outcome of likelihood ratio test of goodness of fit.

```{r model_astar}
# Null model
astar_null <- lmer(a_star ~ 1 + (1 | Animal_ID),
                   data = data,
                   REML = FALSE)

## First-order polynomial (linear) model
astar_linear <- lmer(a_star ~ O2Hb + (1 | Animal_ID),
                     data = data,
                     REML = FALSE)

## Second-order polynomial (quadratic) model
astar_poly <- lmer(a_star ~ poly(O2Hb, 2) + (1 | Animal_ID),
                   data = data,
                   REML = FALSE)

# Compare model fits
anova(astar_null, astar_linear, astar_poly, test = 'LRT')
```

Best fit model is the second-order polynomial (quadratic) model.

## Explore the best fit model

```{r astar_poly}
# Summary output
summary(astar_poly)

# Check model assumptions
plot_model(model = astar_poly,
           type = 'diag')[2:4]

# R-squared (based on likelihood ratio test)
r.squaredLR(astar_poly, null = astar_null)

# Root mean square error
RMSE.merMod(astar_poly)
```

## Generate prediction intervals

```{r astar_interval}
# Prediction interval
astar_lqmm <- lqmm(fixed = a_star ~ poly(O2Hb, 2),
                   random = ~ 1,
                   group = ~ Animal_ID,
                   tau = c(0.025, 0.976),
                   data = data)

astar_interval <- as.data.frame(predict(astar_lqmm, level = 0))

names(astar_interval) <- paste0('Q', c(2.5, 97.5))

astar_interval <- cbind(data, astar_interval)

# Numeric output
astar_interval %>% 
    dplyr::select(O2Hb, a_star, Q2.5, Q97.5) %>% 
    arrange(O2Hb) %>% 
    unique(.) %>% 
    kable(caption = 'a*: 95% prediction interval',
          col.names = c('FO2Hb (%)', 'a*', 'Q2.5', 'Q97.5'))

# Plot output
ggplot(data = astar_interval) + 
    aes(x = O2Hb) +
    geom_point(data = data,
               aes(y = a_star,
                   x = O2Hb,
                   fill = Animal_ID),
               shape = 21,
               size = 3) +
    geom_smooth(data = astar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = astar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs a*',
         subtitle = '95% prediction interval',
         y = 'a*',
         x = 'FO2Hb (%)')
```

## Generate predicted values 

```{r astar_predicted}
# New data
astar_grid <- expand.grid(O2Hb = c(15, 25, 35, 45, 55, 65, 75, 85, 95),
                          Animal_ID = 'I12')

# Generate predicted vs independent variable O2Hb
astar_predicted <- predict(astar_poly, 
                           astar_grid, 
                           allow.new.levels = TRUE)

# Tabulate
data.frame(O2Hb = astar_grid$O2Hb,
           astar = astar_predicted) %>% 
    kable(caption = 'a* predicted values',
          col.names = c('FO2Hb (%)', 'a*'))
```

----

# b*

## Exploratory plots

### Linear model

```{r exploratory_bstarA}
# Composite plot
ggplot(data = data) +
    aes(y = b_star,
        x = O2Hb,
        fill = Animal_ID) +
    geom_point(shape = 21,
               size = 3) +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm',
                fill = 'blue') +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs b*',
         subtitle = 'All animals together\nRegression line is simple linear model',
        x = 'FO2Hb (%)',
        y = 'b*') 
```

```{r exploratory_bstarB, fig.height = 8}
# Small multiples
ggplot(data = data) +
    aes(y = b_star,
        x = O2Hb) +
    geom_point(shape = 21,
               size = 3, 
               fill = '#FFFFFF') +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm') +
    labs(title = 'FO2Hb vs b*',
         subtitle = 'Facetted by animal\nRegression line: linear',
         y = 'b*',
         x = 'FO2Hb (%)') +
    facet_wrap(~ Animal_ID, 
               ncol = 3)
```

### Quadratic model

```{r exploratory_bstarC}
# Composite plot
ggplot(data = data) +
    aes(y = b_star,
        x = O2Hb,
        fill = Animal_ID) +
    geom_point(shape = 21,
               size = 3) +
    geom_smooth(se = FALSE,
                size = 0.8,
                method = 'lm',
                fill = 'blue',
                formula = y ~ poly(x, 2)) +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs b*',
         subtitle = 'All animals together\nRegression line is a quadratic polynomial model',
        x = 'FO2Hb (%)',
        y = 'b*') 
```

```{r exploratory_bstarD, fig.height = 8}
# Small multiples
ggplot(data = data) +
    aes(y = b_star,
        x = O2Hb) +
    geom_point(shape = 21,
               size = 3, 
               fill = '#FFFFFF') +
    geom_smooth(se = FALSE,
                size = 0.8,
                colour = 'blue',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    labs(title = 'FO2Hb vs b*',
         subtitle = 'Facetted by animal\nRegression line: quadratic polynomial',
         y = 'b*',
         x = 'FO2Hb (%)') +
    facet_wrap(~ Animal_ID, 
               ncol = 3)
```

## Model selection

Model selection based on the outcome of likelihood ratio test of goodness of fit.

```{r model_bstar}
# Null model
bstar_null <- lmer(b_star ~ 1 + (1 | Animal_ID),
                   data = data,
                   REML = FALSE)

## First-order polynomial (linear) model
bstar_linear <- lmer(b_star ~ O2Hb + (1 | Animal_ID),
                     data = data,
                     REML = FALSE)

## Second-order polynomial (quadratic) model
bstar_poly <- lmer(b_star ~ poly(O2Hb, 2) + (1 | Animal_ID),
                   data = data,
                   REML = FALSE)

# Compare model fits
anova(bstar_null, bstar_linear, bstar_poly, test = 'LRT')
```

Best fit model is the second-order polynomial (quadratic) model.

## Explore the best fit model

```{r bstar_poly}
# Summary output
summary(bstar_poly)

# Check model assumptions
plot_model(model = bstar_poly,
           type = 'diag')[2:4]

# R-squared (based on likelihood ratio test)
r.squaredLR(bstar_poly, null = bstar_null)

# Root mean square error
RMSE.merMod(bstar_poly)
```

## Generate prediction intervals

```{r bstar_interval}
# Prediction interval
bstar_lqmm <- lqmm(fixed = b_star ~ poly(O2Hb, 2),
                   random = ~ 1,
                   group = ~ Animal_ID,
                   tau = c(0.025, 0.976),
                   data = data)

bstar_interval <- as.data.frame(predict(bstar_lqmm, level = 0))

names(bstar_interval) <- paste0('Q', c(2.5, 97.5))

bstar_interval <- cbind(data, bstar_interval)

# Numeric output
bstar_interval %>% 
    dplyr::select(O2Hb, b_star, Q2.5, Q97.5) %>% 
    arrange(O2Hb) %>% 
    unique(.) %>% 
    kable(caption = 'b*: 95% prediction interval',
          col.names = c('FO2Hb (%)', 'b*', 'Q2.5', 'Q97.5'))

# Plot output
ggplot(data = bstar_interval) + 
    aes(x = O2Hb) +
    geom_point(data = data,
               aes(y = b_star,
                   x = O2Hb,
                   fill = Animal_ID),
               shape = 21,
               size = 3) +
    geom_smooth(data = bstar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = bstar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    scale_fill_viridis_d(name = 'Animal') +
    labs(title = 'FO2Hb vs b*',
         subtitle = '95% prediction interval',
         y = 'b*',
         x = 'FO2Hb (%)')
```

## Generate predicted values 

```{r bstar_predicted}
# New data
bstar_grid <- expand.grid(O2Hb = c(15, 25, 35, 45, 55, 65, 75, 85, 95),
                          Animal_ID = 'I12')

# Generate predicted vs independent variable O2Hb
bstar_predicted <- predict(bstar_poly, 
                           bstar_grid, 
                           allow.new.levels = TRUE)

# Tabulate
data.frame(O2Hb = bstar_grid$O2Hb,
           bstar = bstar_predicted) %>% 
    kable(caption = 'b* predicted values',
          col.names = c('FO2Hb (%)', 'b*'))
```

----

# Code for publication plots

```{r pub_plots}
# Individual plots
l <- ggplot(data = lstar_interval) + 
    aes(x = O2Hb) +
    geom_point(data = data,
               aes(y = l_star),
               size = 3,
               shape = 21,
               fill = '#FFFFFF') +
    geom_smooth(data = data,
                aes(y = l_star),
                se = FALSE,
                colour = '#000000',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    geom_smooth(data = lstar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = lstar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    annotate(geom = 'text', 
             x = 29, y = 24.2,
             label = 'Pseudo-R ^2 == 0.94', 
             parse = TRUE, 
             size = 4) +
    annotate(geom = 'text', 
             x = 25, y = 23.2,
             label = 'RMSE = 0.67', 
             size = 4) +
    scale_y_continuous(limits = c(8, 25),
                       breaks = c(12, 16, 20, 24),
                       expand = c(0, 0)) +
    labs(y = 'L*',
         x = expression('FO'[2]*'Hb (%)'))

ggsave(filename = 'figures/fo2hb-L.png',
       plot = l,
       height = 4,
       width = 5)

a <- ggplot(data = astar_interval) + 
    aes(x = O2Hb) +
    geom_point(data = data,
               aes(y = a_star),
               size = 3,
               shape = 21,
               fill = '#FFFFFF') +
    geom_smooth(data = data,
                aes(y = a_star),
                se = FALSE,
                colour = '#000000',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    geom_smooth(data = astar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = astar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    annotate(geom = 'text', 
             x = 29, y = 52,
             label = 'Pseudo-R ^2 == 0.93', 
             parse = TRUE, 
             size = 4) +
    annotate(geom = 'text', 
             x = 25, y = 50,
             label = 'RMSE = 1.39', 
             size = 4) +
    scale_y_continuous(limits = c(18, 52),
                       breaks = c(20, 30, 40, 50)) +
    labs(y = 'a*',
         x = expression('FO'[2]*'Hb (%)')) 

ggsave(filename = 'figures/fo2hb-a.png',
       plot = a,
       height = 4,
       width = 5)
    
b <- ggplot(data = bstar_interval) + 
    aes(x = O2Hb) +
    geom_point(data = data,
               aes(y = b_star),
               size = 3,
               shape = 21,
               fill = '#FFFFFF') +
    geom_smooth(data = data,
                aes(y = b_star),
                se = FALSE,
                colour = '#000000',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    geom_smooth(data = bstar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = bstar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    annotate(geom = 'text', 
             x = 29, y = 39,
             label = 'Pseudo-R ^2 == 0.91', 
             parse = TRUE, 
             size = 4) +
    annotate(geom = 'text', 
             x = 25, y = 37,
             label = 'RMSE = 1.61', 
             size = 4) +
    scale_y_continuous(limits = c(6, 39),
                       breaks = c(10, 20, 30)) +
    labs(y = 'b*',
         x = expression('FO'[2]*'Hb (%)')) 

ggsave(filename = 'figures/fo2hb-b.png',
       plot = b,
       height = 4,
       width = 5)

# Stacked plot
lplot <- ggplot(data = lstar_interval) + 
    aes(x = O2Hb) +
    geom_smooth(data = data,
                aes(y = l_star),
                se = FALSE,
                colour = '#000000',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    geom_smooth(data = lstar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = lstar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    labs(y = 'L*') +
    theme(axis.title.x = element_blank())

aplot <- ggplot(data = astar_interval) + 
    aes(x = O2Hb) +
    geom_smooth(data = data,
                aes(y = a_star),
                se = FALSE,
                colour = '#000000',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    geom_smooth(data = astar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = astar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    labs(y = 'a*') +
    theme(axis.title.x = element_blank(),
          legend.position = 'none')
    
bplot <- ggplot(data = bstar_interval) + 
    aes(x = O2Hb) +
    geom_smooth(data = data,
                aes(y = b_star),
                se = FALSE,
                colour = '#000000',
                method = 'lm',
                formula = y ~ poly(x, 2, raw = TRUE)) +
    geom_smooth(data = bstar_interval,
                aes(y = `Q2.5`),
                se = FALSE) +
    geom_smooth(data = bstar_interval,
                aes(y = `Q97.5`),
                se = FALSE) +
    labs(y = 'b*',
         x = expression('FO'[2]*'Hb (%)')) +
    theme(legend.position = 'none')

# Construct plot
pub_plot <- lplot + aplot + bplot + 
    plot_layout(ncol = 1) + 
    plot_annotation(tag_levels = 'A')

ggsave(filename = 'figures/figure_1b.png', 
       plot = pub_plot,
       height = 11,
       width = 6)
```
----

# Session Information

```{r session}
sessionInfo()
```
