---
title: 'Supplement 4'
subtitle: "FO2Hb and SaO2 with other co-oximetry covariates"
author: "Peter Kamerman and Etienne Basson"
date: "Last updated: `r format(Sys.Date(), '%d %B %Y')`"
mainfont: Helvetica
geometry: margin=15mm
---

```{r setup, include = FALSE}
# Load packages
library(magrittr)
library(lmerTest)
library(merTools)
library(MuMIn)
library(tidyverse)
library(knitr)

# ggplot theme
theme_set(new = theme_bw(base_size = 18) +
              theme(panel.grid = element_blank(),
                    plot.title = element_text(size = 18),
                    plot.subtitle = element_text(size = 14)))

# Create figures directory
if(!dir.exists('figures')){
    dir.create('figures')
}

# Set chunk options
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      fig.path = 'figures/supplement-4-covariates/',
                      fig.height = 5.8)
```

----

# Research question

1. Determine whether along with functional oxyhaemoglobin saturation (SaO2, %Hb) and fractional oxyhaemoglobin saturation (FO2Hb, %Hb), total haemoglobin concentration (Hb, g/dl), carboxyhaemoglobin (COHb, %Hb), and methaemoglobin (MetHb, %Hb) are significant covariates in the prediction of each CIE L\*a\*b\* component.

----

# Load data 

```{r load_data}
data <- read_rds('data-cleaned/clean-data.rds')
```

----

# Quick look at the data

```{r quick_look}
dim(data)
names(data)
glimpse(data)
```

----

# Prepare data

```{r prepare_data}
# Convert Animal ID to a factor
data %<>%
    mutate(Animal_ID = factor(Animal_ID,
                              levels = c('I1', 'I2', 'I3', 'I4', 'I5',
                                         'I6', 'I8', 'I9', 'I10', 'I11')))
```

----

# FO2Hb

## L*

### Exploratory plots

```{r lstar_plotFO2Hb}
lstar <- data %>% 
    select(Animal_ID, l_star, O2Hb, total_Hb, COHb, MetHb) %>% 
    pivot_longer(cols = c(O2Hb, total_Hb, COHb, MetHb),
                 names_to = 'Variable',
                 values_to = 'Values') %>% 
    mutate(Variable = factor(Variable,
                             levels = c('O2Hb', 'total_Hb', 'COHb', 'MetHb'),
                             labels = c('FO2Hb (%)', 'Hb (g/dl)', 
                                        'COHb (%)', 'MetHb (%)'))) %>% 
    group_by(Variable) %>% 
    nest() %>% 
    mutate(plot = map2(.x = data,
                       .y = Variable,
                       ~ .x %>% 
                           ggplot(data = .) +
                           aes(x = Values,
                               y = l_star) +
                           geom_point(aes(fill = Animal_ID),
                                      shape = 21,
                                      size = 3,
                                      alpha = 0.8) +
                           geom_smooth(se = FALSE,
                                       colour = '#000000') +
                           labs(title = str_glue('L* vs {.y}'),
                                subtitle = 'Black line: LOESS curve',
                                x = .y,
                                y = 'L*') +
                           scale_fill_viridis_d(name = 'Animal')))

walk(lstar$plot, ~print(.x))
```

Appears to be a quadratic fit for COHb, linear fit for MetHb, and no real relationship between L* and [Hb].

### Model

#### Full model

```{r model_lstarFO2Hb1}
# Specify model
lstar_full <- lmer(l_star ~ poly(O2Hb, 2) +
                       poly(COHb, 2) +
                       MetHb +
                       total_Hb + 
                       (1 | Animal_ID),
                   data = data)

# Summarise model
summary(lstar_full) 

# Fixed effect with 95%CI
cbind(fixef(lstar_full), confint(lstar_full)[3:9, ]) %>% 
    kable(caption = 'L* full model: FO2Hb, COHb, MetHb, and [Hb]',
          digits = 3,
          col.names = c('Estimate', '2.5%', '97.5%'))
```

#### Compare model with and without methaemoglobin

```{r model_lstarFO2Hb2}
# FO2Hb only
lstar_FO2Hb <- lmer(l_star ~ poly(O2Hb, 2) +
                        (1 | Animal_ID),
                    data = data)

# FO2Hb + MetHB
lstar_MetHb <- lmer(l_star ~ poly(O2Hb, 2) +
                        MetHb + 
                        (1 | Animal_ID),
                    data = data)

# Compare model fits
anova(lstar_MetHb, lstar_FO2Hb)

# Compare model R-squared values
r.squaredLR(lstar_FO2Hb)
r.squaredLR(lstar_MetHb)

# Compare model RMSE
RMSE.merMod(lstar_FO2Hb)
RMSE.merMod(lstar_MetHb)
```

Likelihood ratio test found model with MetHb had a better fit, but R-squared and RMSE indicate the difference is of marginal importance.

## a*

### Exploratory plots

```{r astar_plotFO2Hb}
astar <- data %>% 
    select(Animal_ID, a_star, O2Hb, total_Hb, COHb, MetHb) %>% 
    pivot_longer(cols = c(O2Hb, total_Hb, COHb, MetHb),
                 names_to = 'Variable',
                 values_to = 'Values') %>% 
    mutate(Variable = factor(Variable,
                             levels = c('O2Hb', 'total_Hb', 'COHb', 'MetHb'),
                             labels = c('FO2Hb (%)', 'Hb (g/dl)', 
                                        'COHb (%)', 'MetHb (%)'))) %>% 
    group_by(Variable) %>% 
    nest() %>% 
    mutate(plot = map2(.x = data,
                       .y = Variable,
                       ~ .x %>% 
                           ggplot(data = .) +
                           aes(x = Values,
                               y = a_star) +
                           geom_point(aes(fill = Animal_ID),
                                      shape = 21,
                                      size = 3,
                                      alpha = 0.8) +
                           geom_smooth(se = FALSE,
                                       colour = '#000000') +
                           labs(title = str_glue('a* vs {.y}'),
                                subtitle = 'Black line: LOESS curve',
                                x = .y,
                                y = 'a*') +
                           scale_fill_viridis_d(name = 'Animal')))

walk(astar$plot, ~print(.x))
```

Appears to be a quadratic fit for COHb, linear fit for MetHb, and no real relationship between L* and [Hb].

### Model

#### Full model

```{r model_astarFO2Hb1}
# Specify model
astar_full <- lmer(a_star ~ poly(O2Hb, 2) +
                       poly(COHb, 2) +
                       MetHb +
                       total_Hb + 
                       (1 | Animal_ID),
                   data = data)

# Summarise model
summary(astar_full) 

# Fixed effect with 95%CI
cbind(fixef(astar_full), confint(astar_full)[3:9, ]) %>% 
    kable(caption = 'a* full model: FO2Hb, COHb, MetHb, and [Hb]',
          digits = 3,
          col.names = c('Estimate', '2.5%', '97.5%'))
```

#### Compare model with and without methaemoglobin and carboxyhaemoglobin

```{r model_astarFO2Hb2}
# FO2Hb only
astar_FO2Hb <- lmer(a_star ~ poly(O2Hb, 2) +
                        (1 | Animal_ID),
                    data = data)

# FO2Hb + MetHB + COHb
astar_MetCOHb <- lmer(a_star ~ poly(O2Hb, 2) +
                        poly(COHb, 2) +
                        MetHb +
                        (1 | Animal_ID),
                    data = data)

# Compare model fits
anova(astar_MetCOHb, astar_FO2Hb)

# Compare model R-squared values
r.squaredLR(astar_FO2Hb)
r.squaredLR(astar_MetCOHb)

# Compare model RMSE
RMSE.merMod(astar_FO2Hb)
RMSE.merMod(astar_MetCOHb)
```

Likelihood ratio test found model with MetHb and COHb had a better fit, but R-squared and RMSE indicate the difference is of marginal importance.

## b*

### Exploratory plots

```{r bstar_plotFO2Hb}
bstar <- data %>% 
    select(Animal_ID, b_star, O2Hb, total_Hb, COHb, MetHb) %>% 
    pivot_longer(cols = c(O2Hb, total_Hb, COHb, MetHb),
                 names_to = 'Variable',
                 values_to = 'Values') %>% 
    mutate(Variable = factor(Variable,
                             levels = c('O2Hb', 'total_Hb', 'COHb', 'MetHb'),
                             labels = c('FO2Hb (%)', 'Hb (g/dl)', 
                                        'COHb (%)', 'MetHb (%)'))) %>% 
    group_by(Variable) %>% 
    nest() %>% 
    mutate(plot = map2(.x = data,
                       .y = Variable,
                       ~ .x %>% 
                           ggplot(data = .) +
                           aes(x = Values,
                               y = b_star) +
                           geom_point(aes(fill = Animal_ID),
                                      shape = 21,
                                      size = 3,
                                      alpha = 0.8) +
                           geom_smooth(se = FALSE,
                                       colour = '#000000') +
                           labs(title = str_glue('b* vs {.y}'),
                                subtitle = 'Black line: LOESS curve',
                                x = .y,
                                y = 'b*') +
                           scale_fill_viridis_d(name = 'Animal')))

walk(bstar$plot, ~print(.x))
```

Appears to be a quadratic fit for COHb, linear fit for MetHb, and no real relationship between L* and [Hb].

### Model

#### Full model

```{r model_bstarFO2Hb1}
# Specify model
bstar_full <- lmer(b_star ~ poly(O2Hb, 2) +
                       poly(COHb, 2) +
                       MetHb +
                       total_Hb + 
                       (1 | Animal_ID),
                   data = data)

# Summarise model
summary(bstar_full) 

# Fixed effect with 95%CI
cbind(fixef(bstar_full), confint(bstar_full)[3:9, ]) %>% 
    kable(caption = 'a* full model: FO2Hb, COHb, MetHb, and [Hb]',
          digits = 3,
          col.names = c('Estimate', '2.5%', '97.5%'))
```

#### Compare model with and without methaemoglobin and carboxyhaemoglobin

```{r model_bstarFO2Hb2}
# FO2Hb only
bstar_FO2Hb <- lmer(b_star ~ poly(O2Hb, 2) +
                        (1 | Animal_ID),
                    data = data)

# FO2Hb + MetHB + COHb
bstar_MetCOHb <- lmer(b_star ~ poly(O2Hb, 2) +
                        poly(COHb, 2) +
                        MetHb +
                        (1 | Animal_ID),
                    data = data)

# Compare model fits
anova(bstar_MetCOHb, bstar_FO2Hb)

# Compare model R-squared values
r.squaredLR(bstar_FO2Hb)
r.squaredLR(bstar_MetCOHb)

# Compare model RMSE
RMSE.merMod(bstar_FO2Hb)
RMSE.merMod(bstar_MetCOHb)
```

Likelihood ratio test found model with MetHb and COHb had a better fit, but R-squared and RMSE indicate the difference is of marginal importance.

----

# SaO2

## L*

### Exploratory plots

```{r lstar_plotSaO2}
lstar <- data %>% 
    select(Animal_ID, l_star, Coox_SaO2, total_Hb, COHb, MetHb) %>% 
    pivot_longer(cols = c(Coox_SaO2, total_Hb, COHb, MetHb),
                 names_to = 'Variable',
                 values_to = 'Values') %>% 
    mutate(Variable = factor(Variable,
                             levels = c('Coox_SaO2', 'total_Hb', 'COHb', 'MetHb'),
                             labels = c('SaO2 (%)', 'Hb (g/dl)', 
                                        'COHb (%)', 'MetHb (%)'))) %>% 
    group_by(Variable) %>% 
    nest() %>% 
    mutate(plot = map2(.x = data,
                       .y = Variable,
                       ~ .x %>% 
                           ggplot(data = .) +
                           aes(x = Values,
                               y = l_star) +
                           geom_point(aes(fill = Animal_ID),
                                      shape = 21,
                                      size = 3,
                                      alpha = 0.8) +
                           geom_smooth(se = FALSE,
                                       colour = '#000000') +
                           labs(title = str_glue('L* vs {.y}'),
                                subtitle = 'Black line: LOESS curve',
                                x = .y,
                                y = 'L*') +
                           scale_fill_viridis_d(name = 'Animal')))

walk(lstar$plot, ~print(.x))
```

Appears to be a quadratic fit for COHb, linear fit for MetHb, and no real relationship between L* and [Hb].

### Model

#### Full model

```{r model_lstarSaO21}
# Specify model
lstar_full <- lmer(l_star ~ poly(Coox_SaO2, 2) +
                       poly(COHb, 2) +
                       MetHb +
                       total_Hb + 
                       (1 | Animal_ID),
                   data = data)

# Summarise model
summary(lstar_full) 

# Fixed effect with 95%CI
cbind(fixef(lstar_full), confint(lstar_full)[3:9, ]) %>% 
    kable(caption = 'L* full model: SaO2, COHb, MetHb, and [Hb]',
          digits = 3,
          col.names = c('Estimate', '2.5%', '97.5%'))
```

#### Compare model with and without methaemoglobin

```{r model_lstarSaO22}
# SaO2 only
lstar_SaO2 <- lmer(l_star ~ poly(Coox_SaO2, 2) +
                        (1 | Animal_ID),
                    data = data)

# SaO2 + MetHB
lstar_MetHb <- lmer(l_star ~ poly(Coox_SaO2, 2) +
                        MetHb + 
                        (1 | Animal_ID),
                    data = data)

# Compare model fits
anova(lstar_MetHb, lstar_SaO2)

# Compare model R-squared values
r.squaredLR(lstar_SaO2)
r.squaredLR(lstar_MetHb)

# Compare model RMSE
RMSE.merMod(lstar_SaO2)
RMSE.merMod(lstar_MetHb)
```

Likelihood ratio test found model with MetHb had a better fit, but R-squared and RMSE indicate the difference is of marginal importance.

## a*

### Exploratory plots

```{r astar_plotSaO2}
astar <- data %>% 
    select(Animal_ID, a_star, Coox_SaO2, total_Hb, COHb, MetHb) %>% 
    pivot_longer(cols = c(Coox_SaO2, total_Hb, COHb, MetHb),
                 names_to = 'Variable',
                 values_to = 'Values') %>% 
    mutate(Variable = factor(Variable,
                             levels = c('Coox_SaO2', 'total_Hb', 'COHb', 'MetHb'),
                             labels = c('SaO2 (%)', 'Hb (g/dl)', 
                                        'COHb (%)', 'MetHb (%)'))) %>% 
    group_by(Variable) %>% 
    nest() %>% 
    mutate(plot = map2(.x = data,
                       .y = Variable,
                       ~ .x %>% 
                           ggplot(data = .) +
                           aes(x = Values,
                               y = a_star) +
                           geom_point(aes(fill = Animal_ID),
                                      shape = 21,
                                      size = 3,
                                      alpha = 0.8) +
                           geom_smooth(se = FALSE,
                                       colour = '#000000') +
                           labs(title = str_glue('a* vs {.y}'),
                                subtitle = 'Black line: LOESS curve',
                                x = .y,
                                y = 'a*') +
                           scale_fill_viridis_d(name = 'Animal')))

walk(astar$plot, ~print(.x))
```

Appears to be a quadratic fit for COHb, linear fit for MetHb, and no real relationship between L* and [Hb].

### Model

#### Full model

```{r model_astarSaO21}
# Specify model
astar_full <- lmer(a_star ~ poly(Coox_SaO2, 2) +
                       poly(COHb, 2) +
                       MetHb +
                       total_Hb + 
                       (1 | Animal_ID),
                   data = data)

# Summarise model
summary(astar_full) 

# Fixed effect with 95%CI
cbind(fixef(astar_full), confint(astar_full)[3:9, ]) %>% 
    kable(caption = 'a* full model: SaO2, COHb, MetHb, and [Hb]',
          digits = 3,
          col.names = c('Estimate', '2.5%', '97.5%'))
```

#### Compare model with and without methaemoglobin and carboxyhaemoglobin

```{r model_astarSaO22}
# SaO2 only
astar_SaO2 <- lmer(a_star ~ poly(Coox_SaO2, 2) +
                        (1 | Animal_ID),
                    data = data)

# SaO2 + MetHB + COHb + [Hb]
astar_MetCOHb <- lmer(a_star ~ poly(Coox_SaO2, 2) +
                          poly(COHb, 2) +
                          MetHb +
                          total_Hb +
                          (1 | Animal_ID),
                      data = data)

# Compare model fits
anova(astar_MetCOHb, astar_SaO2)

# Compare model R-squared values
r.squaredLR(astar_SaO2)
r.squaredLR(astar_MetCOHb)

# Compare model RMSE
RMSE.merMod(astar_SaO2)
RMSE.merMod(astar_MetCOHb)
```

Likelihood ratio test found model with MetHb, COHb and [Hb] had a better fit, but R-squared and RMSE indicate the difference is of marginal importance.

## b*

### Exploratory plots

```{r bstar_plotSaO2}
bstar <- data %>% 
    select(Animal_ID, b_star, Coox_SaO2, total_Hb, COHb, MetHb) %>% 
    pivot_longer(cols = c(Coox_SaO2, total_Hb, COHb, MetHb),
                 names_to = 'Variable',
                 values_to = 'Values') %>% 
    mutate(Variable = factor(Variable,
                             levels = c('Coox_SaO2', 'total_Hb', 'COHb', 'MetHb'),
                             labels = c('SaO2 (%)', 'Hb (g/dl)', 
                                        'COHb (%)', 'MetHb (%)'))) %>% 
    group_by(Variable) %>% 
    nest() %>% 
    mutate(plot = map2(.x = data,
                       .y = Variable,
                       ~ .x %>% 
                           ggplot(data = .) +
                           aes(x = Values,
                               y = b_star) +
                           geom_point(aes(fill = Animal_ID),
                                      shape = 21,
                                      size = 3,
                                      alpha = 0.8) +
                           geom_smooth(se = FALSE,
                                       colour = '#000000') +
                           labs(title = str_glue('b* vs {.y}'),
                                subtitle = 'Black line: LOESS curve',
                                x = .y,
                                y = 'b*') +
                           scale_fill_viridis_d(name = 'Animal')))

walk(bstar$plot, ~print(.x))
```

Appears to be a quadratic fit for COHb, linear fit for MetHb, and no real relationship between L* and [Hb].

### Model

#### Full model

```{r model_bstarSaO21}
# Specify model
bstar_full <- lmer(b_star ~ poly(Coox_SaO2, 2) +
                       poly(COHb, 2) +
                       MetHb +
                       total_Hb + 
                       (1 | Animal_ID),
                   data = data)

# Summarise model
summary(bstar_full) 

# Fixed effect with 95%CI
cbind(fixef(bstar_full), confint(bstar_full)[3:9, ]) %>% 
    kable(caption = 'a* full model: SaO2, COHb, MetHb, and [Hb]',
          digits = 3,
          col.names = c('Estimate', '2.5%', '97.5%'))
```

#### Compare model with and without methaemoglobin and carboxyhaemoglobin

```{r model_bstarSaO22}
# SaO2 only
bstar_SaO2 <- lmer(b_star ~ poly(Coox_SaO2, 2) +
                        (1 | Animal_ID),
                    data = data)

# SaO2 + MetHB + COHb
bstar_MetCOHb <- lmer(b_star ~ poly(Coox_SaO2, 2) +
                        poly(COHb, 2) +
                        MetHb +
                        (1 | Animal_ID),
                    data = data)

# Compare model fits
anova(bstar_MetCOHb, bstar_SaO2)

# Compare model R-squared values
r.squaredLR(bstar_SaO2)
r.squaredLR(bstar_MetCOHb)

# Compare model RMSE
RMSE.merMod(bstar_SaO2)
RMSE.merMod(bstar_MetCOHb)
```

Likelihood ratio test found model with MetHb and COHb had a better fit, but R-squared and RMSE indicate the difference is of marginal importance.

----

# Session Information

```{r session}
sessionInfo()
```
